Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-07 Manuscript/manuscript.Rmd
updates

In this study we have developed an ensemble framework that aims to address and compare a variety of data irregularities and characteristics that are associated with the prediction of TAD boundaries. We evaluated domain data obtained from four different resolution contact matrices for the GM12878 cell line including 5kb, 25kb, 50kb, and 100kb. For each resolution, we examined the inclusion of three different predictor types as the feature space for downstream modeling. We coupled each predictor type with two separate variable reduction techniques. Lastly, we performed a series of three different re-sampling techniques aimed at creating balanced classes. Random forest classification algorithms were performed to both compare each combination of the above ensemble, and investigate which functional genomic features were most associated with the formation of TAD boundaries.  Model performance was evaluated based on a total of four metrics which included test accuracy, area under the ROC curve (AUC), F1 score, and Matthew's Correlation Coefficient (MCC).

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution (Figure 1). For example, for a 5kb resolution contact matrix, the genome was binned into 5kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sized by regions of size of the resolution of the corresponding contact matrix. For example, for the domain data obtained from a 5kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 5kb for a total width around each boundary point of 10kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

2. **Overlap Percent (OP)**: For each genomic bin, the percent overlap between the feature width and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with the genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 5kb, 25kb, 50kb, or 100kb). Values for bins with no overlaps were set to be 0.

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique) [@chawla2002smote]: This method incorporated both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class. The user is able to specify the parameters denoting the percent of over- and under-sampling from the two classes. We used SMOTE to create perfectly balanced data. The `DMwR` R package (version 3.4.2) was used to perform SMOTE.

@@

We evaluated and compared Hi-C datasets at four different resolutions R = {5kb, 25kb, 50kb, 100kb}, used three predictor types P = {OC, OP, Distance}, two variable reduction techniques V = {LASSO, Elastic-Net}, and four class balancing techniques C = {None, RUS, ROS, SMOTE}. Therefore, a system produced from the ensemble framework is defined as: E = {r, p, v, c}, where r $\in$ R, p $\in$ P, v $\in$ V, and c $\in$ C. Letting |X| denote the cardinality of each set, we evaluated |R| x |P| x |V| x |C| different ensemble systems, totaling 96 (=4 $\times$ 3 $\times$ 2 $\times$ 4) separate models. For simplicity, we present the most significant results, with the complete results available as Supplementary Tables.

@@

The coordinates of the corresponding 5kb resolution contact maps resulted in a total of 44948 genomic bins [??? To be modified for the whole genome] (Table 2). There were only 3083 (6.9%) bins that overlapped with a flanked TAD boundary [??? What is flanked TAD boundary? Why do we care?]. An increase in resolution resulted in an overall decrease in the number of genomic bins, as well as a slighly less degree of class imbalance. The percentage of imbalance for each of the other resolutions were found to be  14.0%, 13.7%, and 12.2% for 25 kb, 50kb, and 100kb resolutions respectively. [??? Add exact numbers of class imbalance for each resolution; !!! added exact numbers] (Supplementaty Table 2). [??? Do not understand. I understand the ratio of #minority/#majority, but what % tells?; !!!was overally average of class imbalance for the 4 data resolutions; removed sentence; doesn't seem relevant ??? We should rewrite using the ratio of #minority/#majority].  After applying each resampling technique to the training sets for each data resolution, we obtained perfectly balanced classes (Figure 4). That is, the ratio of bins that overlapped with flanked TAD boundaries to bins that did not was 1:1.

@@

When comparing re-sampling techniques it was found that performing 50 iterations of random under-sampling yielded significantly higher results for both F1-Scores and MCC metrics (Figure 6).

Random under-sampling substantially improved classification performance as quantified by F1 and MCC metrics (Figure 14). The increase in performance was present across both overlap count and overlap percent predictor types [??? Again, where is distance?]. Models also featured a decrease in both accuracy and AUC, which was attributed to less inflated false negative rates (Supplementary Table 11).

Variable reduction was found not to have an affect on resampling, and subsequent performances, as any differences were again marginal [??? Where is the proof?]. Notably, random under-sampling improved performance when using data at the other resolutions; however, at lower resolutions the performances were lower (Supplementary Figure 15; Supplementary Table 12). These results demonstrate that random under-sampling effectively alleviates class imbalance problem, especially for TAD boundaries detected from high-resolution Hi-C data.

@@

Diagram of the model construction used for downstream analysis. The linear genome was binned according to the resolution of the respective HiC experiment (either 5kb, 25kb, 50kb, or 100kb intervals). TAD boundaries were then flanked by 1-unit on either side of the boundary point. The unit flanking was indicative of the resolution that the domain data was obtained from (i.e. for 5kb resolution, 1 unit represents 5kb for a total flanking region of 10kb). The response vector Y used for classification was determined by whether or not a genomic bin overlapped with a flanked region. The positional coordinates of each functional genomic element, obtained from ENCODE, were then used to define the feature space of the models.

@@

>>>>

In this study we have developed an ensemble framework that aims to address and compare a variety of data irregularities and characteristics that are associated with the prediction of TAD boundaries. We evaluated domain data obtained from four different resolution contact matrices for the GM12878 cell line including 10 kb, 25 kb, 50 kb, and 100 kb. For each resolution, we examined the inclusion of three different predictor types as the feature space for downstream modeling. We coupled each predictor type with two separate variable reduction techniques. Lastly, we performed a series of three different re-sampling techniques aimed at creating balanced classes. Random forest classification algorithms were performed to both compare each combination of the above ensemble, and investigate which functional genomic features were most associated with the formation of TAD boundaries.  Model performance was evaluated based on a total of four metrics which included test accuracy, area under the ROC curve (AUC), F1 score, and Matthew's Correlation Coefficient (MCC).

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution (Figure 1). For example, for a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sized by regions of size of the resolution of the corresponding contact matrix. For example, for the domain data obtained from a 10 kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 10 kb for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

2. **Overlap Percent (OP)**: For each genomic bin, the percent overlap between the feature width and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with the genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 10 kb, 25 kb, 50 kb, or 100 kb). Values for bins with no overlaps were set to be 0.

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique) [@chawla2002smote]: This method incorporated both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE The user is able to specify the parameters denoting the percent of over- and under-sampling from the two classes. We chose the parameters that would create perfectly balanced classes. The `DMwR` R package (version 3.4.2) was used to perform SMOTE.

@@

We evaluated and compared Hi-C datasets at four different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}, used three predictor types P = {OC, OP, Distance}, two variable reduction techniques V = {LASSO, Elastic-Net}, and four class balancing techniques C = {None, RUS, ROS, SMOTE}. Therefore, a system produced from the ensemble framework is defined as: E = {r, p, v, c}, where r $\in$ R, p $\in$ P, v $\in$ V, and c $\in$ C. Letting |X| denote the cardinality of each set, we evaluated |R| x |P| x |V| x |C| different ensemble systems, totaling 96 (=4 $\times$ 3 $\times$ 2 $\times$ 4) separate models. For simplicity, we present the most significant results, with the complete results available as Supplementary Tables.

@@

The coordinates of the corresponding 10 kb resolution contact maps resulted in a total of 44948 genomic bins [??? To be modified for the whole genome] (Table 2). There were only 3083 (6.9%) bins that overlapped with a flanked TAD boundary [??? What is flanked TAD boundary? Why do we care?]. An increase in resolution resulted in an overall decrease in the number of genomic bins, as well as a slighly less degree of class imbalance. The percentage of imbalance for each of the other resolutions were found to be  14.0%, 13.7%, and 12.2% for 25 kb, 50kb, and 100kb resolutions respectively. [??? Add exact numbers of class imbalance for each resolution; !!! added exact numbers] (Supplementaty Table 2). [??? Do not understand. I understand the ratio of #minority/#majority, but what % tells?; !!!was overally average of class imbalance for the 4 data resolutions; removed sentence; doesn't seem relevant ??? We should rewrite using the ratio of #minority/#majority].  After applying each resampling technique to the training sets for each data resolution, we obtained perfectly balanced classes (Figure 4). That is, the ratio of bins that overlapped with flanked TAD boundaries to bins that did not was 1:1.

@@

When comparing re-sampling techniques it was found that performing 50 iterations of random under-sampling yielded the best performances for both F1-Scores and MCC metrics, across all resolutions and predictor types (Figure 6). Models using 50-iterations of the SMOTE algorithm performed comparitively similar to those using RUS, across all resolutions and predictor types (supplementary Figure 1A). However, it was observed that models using 50-iterations of random over-sampling performed substantially worse than both RUS and SMOTE (Supplementary Figure 1B). More importantly, there was a notable decrease in both accuracy and AUC metrics when performing random under-sampling compared to models with no class balancing. This was evidenced by the more balanced interplay between the componenets of the confusion matrices created when validating the models (Supplementary Table 4). That is, less weight was carried by the true negative cases. This feature was also observed for models using SMOTE and ROS (Supplementary Table 5). As was the case for models with no class-balancing, there was a clear improvement when performing RUS for models with distance-type predictors, across all resolutions.[??? Again, where is distance?; !!!included]

@@

Diagram of the model construction used for downstream analysis. The linear genome was binned according to the resolution of the respective HiC experiment (either 10 kb, 25 kb, 50 kb, or 100 kb intervals). TAD boundaries were then flanked by 1-unit on either side of the boundary point. The unit flanking was indicative of the resolution that the domain data was obtained from (i.e. for 10 kb resolution, 1 unit represents 10 kb for a total flanking region of 20 kb). The response vector Y used for classification was determined by whether or not a genomic bin overlapped with a flanked region. The positional coordinates of each functional genomic element, obtained from ENCODE, were then used to define the feature space of the models.

@@

# Figure 6:
